P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2024-06-28 manuscript.tex
Improve theoretical background formulas
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

A \ac{MD} system to be simulated constitutes of a number of atoms, the initial state (such as the atoms' position or velocities), and the boundary conditions of the system.
Here, we use \ac{PBC} in all directions, hence when one atom crosses the domain, it reappears on the opposite side with the same velocity.
%These simulations are defined by establishing the number of atoms in a system, the initial state of the system (such as the atoms' positions or velocities), and the boundary conditions for the system.
Fundamentally, atom trajectories in classical \ac{MD} systems are computed by integrating Newton-Euler equations for every atom $i$:
\begin{equation}
    \hat F_i = m \dot{\hat v}_i \label{eq:newton_force}
\end{equation}
\begin{equation}
    \hat v_i = \dot{\hat x}_i \label{eq:newton_velocity}
\end{equation}
The forces $F_i$ are described by the negative gradient of the potential of interest.
The mutual force between atom $i$ and $j$ as governed by the \ac{LJ} potential
is given as:
\begin{equation}
    \hat{F}_{LJ}(\hat{x_i}, \hat{x_j}) = -\frac{24\varepsilon}{x_{ij}} \left( \frac{\sigma}{x_{ij}} \right)^{6} \left[ 2\left(\frac{\sigma}{x_{ij}}\right)^{6} - 1\right] \hat{x}_{ij}~,
    %V_{LJ} = 4\varepsilon \left[ \left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right]
    \label{eq:lennard_jones}
\end{equation}
Where $\hat{x}_{ij}$ is the distance vector between atoms $i$ and $j$, $x_{ij}$ is the distance between atoms $i$ and $j$, $\varepsilon$ is the width of the potential well, and $\sigma$ specifies at which distance the potential is~$0$.

@@

\begin{equation}
    \hat{\tau}{_i} = I \dot{\hat \omega}_i + {\hat \omega}_i \times I {\hat \omega}_i \label{eq:newton_torque}
\end{equation}
\begin{equation}
    \hat{\omega}{_i} = \dot{\hat{\varphi}}_i \label{eq:newton_angular_velocity}
\end{equation}

Also, a collision detection step is also needed to check when two particles overlap.
When particles overlap, a collision is detected and information regarding the collision such as the contact point and penetration depth are calculated.
Note that there may be penetration when a collision happens because there is no guarantee that it is detect right when it starts, as the simulation is discretized by the time-step size.
For simplification purposes, we only calculate collision detections analytically in this work since there are no particles with complex shapes involved.

In this work, we use the penalty-based Linear Spring-Dashpot model to calculate forces among particles for the DEM cases.
The collision force is split into two components, the force in the normal direction and in the tangential direction.

@@

Where $k^{n}$ is the spring stiffness, $\xi^{n}$ the normal velocity damping and $\mu$ the friction.
The update rule \autoref{eq:update_rule} and modifications in \autoref{eq:linear_spring_dashpot_tangential} were properly included to fix frictional contact accordingly \cite{dem1,dem2}.
The update rule is carried out every time-step the contact persists, which from the algorithmic point of view means that a contact history data structure is needed to properly compute these interactions.

@@

To obtain the force for a particle in a specific time-step, it is necessary to compute its partial contributions from each of the other particles present in the system.
In a naive strategy, this results in a quadratic complexity and severely affects the application performance, which limits the system sizes and time-scales of such simulations.

>>>>

A \ac{MD} system to be simulated constitutes of a number of atoms, the initial state (such as the atoms' position or velocities), and the boundary conditions of the system.
Here, we use \ac{PBC} in all directions, hence when one atom crosses the domain, it reappears on the opposite side with the same velocity.
%These simulations are defined by establishing the number of atoms in a system, the initial state of the system (such as the atoms' positions or velocities), and the boundary conditions for the system.
Fundamentally, atom trajectories in classical \ac{MD} systems are computed by integrating Newton-Euler equations for every atom $i$:
\begin{equation}
    \hat F_i = m_i \dot{\hat v}_i \label{eq:newton_force}
\end{equation}
\begin{equation}
    \hat v_i = \dot{\hat x}_i \label{eq:newton_velocity}
\end{equation}
Where $m_i$, $\hat v_i$ and $\hat x_i$ are the mass, the velocity vector and the position vector of the atom $i$, respectively.
The force $F_i$ then must be calculated to integrate the positions of the atoms, and these are described by the negative gradient of the potential of interest.
The mutual force between atom $i$ and $j$ as governed by the \ac{LJ} potential
is given as:
\begin{equation}
    \hat{F}_{LJ}(\hat{x_i}, \hat{x_j}) = -\frac{24\varepsilon}{x_{ij}} \left( \frac{\sigma}{x_{ij}} \right)^{6} \left[ 2\left(\frac{\sigma}{x_{ij}}\right)^{6} - 1\right] \hat{x}_{ij}~,
    %V_{LJ} = 4\varepsilon \left[ \left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right]
    \label{eq:lennard_jones}
\end{equation}
Where $\hat{x}_{ij}$ is the distance vector between atoms $i$ and $j$, $x_{ij}$ is the distance between atoms $i$ and $j$, $\varepsilon$ is the width of the potential well, and $\sigma$ specifies at which distance the potential is~$0$.

@@

\begin{equation}
    \hat{\tau}{_i} = I \dot{\hat \omega}_i + {\hat \omega}_i \times I {\hat \omega}_i \label{eq:newton_torque}
\end{equation}
\begin{equation}
    \hat{\omega}{_i} = \dot{\varphi}_i \label{eq:newton_angular_velocity}
\end{equation}

With the torque vector $\hat{\tau}{_i}$, angular velocity vector $\hat{\omega}{_i}$, rotation $\varphi$ (typically stored as a quaternion, rotation matrix or rotation vector) and moment of inertia about the center of mass $I = I(\varphi_i)$.
A collision detection step is required to check when two particles overlap, and parameters about the collision such as the contact point and penetration depth are calculated and used in the force calculation.
Note that the reason a penetration occurs during a collision is that the collision is not detected right when it starts because the simulation is discretized by the time-step size.
For simplification purposes, we only calculate collision detections analytically in this work since there are no particles with complex shapes involved.

In this work, we use the penalty-based Linear Spring-Dashpot model to calculate forces among particles for the DEM cases.
The collision force is split into two components, one representing the force in the normal direction and one representing the force in the tangential direction.

The relative contact velocity $\hat{u}_{ij}$ is of importance for most contact models.
Considering $\hat{c}_{ij}$ the contact point and $\hat{n}_{ij}$ the normal direction vector between the particles $i$ and $j$, the relative contact velocity and their normal and tangential components can be expressed as:

\begin{equation}
    \hat{u}_{ij} = \hat{v}_{i} - \hat{v}_{j} + \hat{\omega}_{i} \times (\hat{c}_{ij} - \hat{x}_{i}) - \hat{\omega}_{j} \times (\hat{c}_{ij} - \hat{x}_{j})
    \label{eq:relative_contact_velocity}
\end{equation}

\begin{equation}
    \hat{u}_{ij}^{n} = \hat{u}_{ij} \cdot \hat{n}_{ij} * \hat{n}_{ij}
    \label{eq:relative_contact_velocity_n}
\end{equation}

\begin{equation}
    \hat{u}_{ij}^{t} = \hat{u}_{ij} - \hat{u}_{ij}^{n}
    \label{eq:relative_contact_velocity_t}
\end{equation}

Then, the force components in the normal direction $\hat{F}_{SD}^{n}$ and in the tangential direction $\hat{F}_{SD}^{t}$ can be defined as:

@@

Where $k^{n}$ is the spring stiffness, $\xi^{n}$ the normal velocity damping, $\mu$ the friction, and $\delta_{ij}$ the penetration depth.
The update rule \autoref{eq:update_rule} and modifications in \autoref{eq:linear_spring_dashpot_tangential} were properly included to fix frictional contact accordingly \cite{dem1,dem2}.
The update rule is carried out every time-step the contact persists, which from the algorithmic point of view means that a contact history data structure is needed to properly compute these interactions.

@@

To calculate the resulting force for a particle in a specific time-step, it is necessary to compute all partial force contributions from each of the other particles present in the system.
In a naive strategy, this results in a quadratic complexity and severely affects the application performance, which limits the system size and the time-scale of the simulation.
